Evolution of (dynamic) capabilities and its performance implications: An empirical test and extension

### 2019-09-05 article4_tommylars.rnw
Some focus on the mechanism and mediation

\section{Introduction}
There is ongoing interest in (dynamic) capabilities and how they influence firm
performance \citep{Fainshmidt2016a,Teece2014}. Indeed, \emph{dynamic capabilities} (DC),
and its front runner, the resource-based view \citep{Barney2001,Helfat2007}, is now
considered one of the major theories in strategic management
\citep{Schilke2018}. Arguably, the key reason is that this theory clearly articulates how
firms can alter their operating routines through dynamic capabilities. Thus, a key
ingredient in this theory is that DCs can alter firms operating routines, with subsequent
performance implications.

An issue that has received far less attention, however, is whether dynamics capabilities
themselves undergo an evolution \citep{Schilke2018}. Moreover, if DCs undergo an
evolution, is the evolution inert and past-dependent, or more deliberate and represent
purposeful change in firms DCs over time? Unpacking this is crucial to the development of
DC theory as it goes into the very nature of the issue as to whether and to what extent
DCs can purposefully alter firms operating routines, and the performance implications
stemming from such change \citep{Helfat2007}. It is therefore surprising that the
literature has dealt with it sparingly. What is the exception to this is \cite{Zollo2002a}
highly influential model over DC evolution where they postulate that (dynamic)
capabilities evolve with regard to firms’ \emph{deliberate learning} (DL). However, Zollo
\& Winter`s insightful theorizing has received scant empirical testing. Moreover, their
model can be developed further, by incorporating insights from subsequent research and
theorizing since this model was published. One salient issue in this regard is that Zollo
\& Winter did not consider how DCs can influence firms’ DL. Moreover, their model does not
consider how firm performance may be influenced by the evolution of (dynamic)\todo{should
  we really go with the parantheses around dynamic throuhgt the whole paper - it feels
  like we are not certain about the concept} capabilities (through DL). Such feedback
loops may significantly alter how we understand the relationship between deliberate
learning and evolution of (dynamic) capabilities in extant DC theory, including the
performance implications of such evolution. Reflecting this gap in our understanding, we
ask the following research question:

\emph{What is the role of deliberate learning in the evolution of (dynamic) capabilities and the
performance implications for such evolution?}

To answer this research question, we not only empirically test ZW’s a highly influential
model of the evolution of (dynamic) capabilities, but also extend it by incorporating (a)
feedback loops between DCs and deliberate learning and (b) incorporating firm performance
into the model.

\section{Theory}\label{sec:theory}
Firm capabilities can roughly be divided into two categories. First, ordinary capabilities
(OR) “are directed toward maintaining and leveraging the status quo in terms of the scale
and scope of activities, businesses, product lines, customer segments”
\citep[p. 393]{Schilke2018}. In other words these capabilities are the ones who “permit a
firm to ‘make a living’ in the short term” \citep[p. 991]{Winter2003}. On the other hand,
dynamic capabilities (DC) are “those that operate to extend, modify or create ordinary
capabilities” (ibid). DC theory is emerging as an important paradigm in strategic
management research, but also facing some considerable challenges with respect to its
antecedents and outcomes \citep{Eriksson2014,Schilke2018}, as well as its evolution
and development trajectory \citep{Romme2010,Zollo2002a}. Extant
literature has dealt with antecedents in many forms such as ### (ref) etc. Similarly,
outcomes have been the focal point of several contributions focusing on ### (ref)
etc. However, despite DC building on an evolutionary view of resource development, less
attention has been dealt to the evolution of DC over time than one could expect. Partly
this could be due to lacking data (ref), unclear conceptual definitions (ref) or sheer
complexity of studying evolution (ref). One remarkable exception is Zollo \& Winter (2002)
who propose a clear evolutionary argument and hypothesis on DC evolution. Still, little
empirical work has attempted to test their propositions.

\textcolor{red}{DL definition – makes sure all definitions are in order}

By combining insights from 20 years of DC research on external factors such as antecedents
and outcomes with the internal process shaping DC over time, we can shed light on the
workings of DC and extend the knowledge of its implications. Using Zollo \& Winter as our
point of departure we will be able to study the internal process of DC evolution, and by
adding new dimensions related to antecedents and outcomes we will gain insight into
important implications of the concept.

\subsection{Deliberate learning as antecedent and evolution driver}
When Zollo \& Winter (2002) put deliberate learning at the center of their model for DC
evolution they tap into a core tenant of DC theory more developed in later
contributions. Learning theory in organizations distinguishes between autonomous learning
processes \citep{Arthur2005} (e.g. learning by doing \citep{Zollo2002a,Arrow1962} based on
repetitive actions by employees in firms and organizations, and deliberate learning where
learning takes the form of a planned and deliberate process \citep{Arthur2005}. This
deliberation and intent as a driver behind the learning process, corresponds well with the
idea of purposefulness behind strategic intent as evident in the definition of dynamic
capabilities \citep{Helfat2007}: “the ability to {\bf purposefully} build, extend and
modify ordinary capabilities” (our emphasis added). (ref). When organizations invest in
systems to facilitate learning, they invest in capabilities for changing routines in line
with strategic direction of the organization itself. Hence, investments in deliberate
learning corresponds with investments in dynamic capabilities and both concepts contribute
to the development of operational routines.

\emph{H1: Deliberate learning has a direct positive effect on organizations development of
  operating routines.}

They depart conceptually, however, when we look at the functioning of each. The main
function of DL is to accumulate knowledge and experience about the operation of the
organization and its environment (ref). Through this process DL generates insights into
what should be changed and not in order to achieve improved organizational
performance. Contrary, DC deals with how change in operational routines should be done
through putting experience from DL to work (ref). Thus, for DC to be well functioning in
its quest to change OR, a DL process needs to be in place. Without a proper DL the
learning will be less purposeful and hence less attune to the strategic intent of the
firm. Change may still appear, but as organizational change rather than strategic (helfat
distinction between strategix and organizational change). In other words, without DL firms
can achieve technical fitness (ref), but only through DL shaping DC can firms achieve
evolutionary fitness (ref). Consequently, DL is an important antecedent to DC.

\emph{H2: Deliberate learning leads to stronger dynamic capabilities and has a positive
  influence on the level of dynamic capabilities within the firm.}

\emph{H3: Deliberate learning has an indirect effect on operational routines through
  dynamic capabilities.}

Besides impacting the level of DC as an antecedent, DL also contributes to the evolution
of DC over time. Conventional insights suggest that DC cannot be aquired in factor markets
and thus has to be developed internally \citep{Coen2011,Schilke2014a}\todo{(Helfat,
  Finkelstein, Mitchell, Peteraf, Singh,\& Teece, 2007; Katkalo, Pitelis, \& Teece, 2010;
  Maritan \& Peteraf, 2011; Teece, 2014). in Schilke 2014} . This has performance
implications related to the VRIO principle \citep{Barney1991} (as discussed later in this
article), as well as consequences for the evolution of DC. Considering DC as itself a
hierarchy of capabilities \citep{Collis1994,Winter2003} opens up for a distinction between
first-order and second-order DC. On the one hand, systematic learning generates insights
into how routines and lower level capabilities should change (i.e. impacting first-order
DC) \citep{Mahoney1995,Schilke2014a}. On the other, the same learning process is very much
involved in developing and shaping these capabilities over time (Kianto and Ritala 2010;
Schilke2014). Consequently, in addition to being an antecedent to DC, DL also would work
as a second-order DC influencing the evolution of DC over time (the ability of
“learning-to-learn” (Collis, 1994)).

\emph{H4: Deliberate learning has a direct positive effect on changes in dynamic
  capabilities over time}

Dynamic capabilities enables firms to adapt better to changing environments
\citep{Teece1997} and hence creating strategic change in terms of purposefully changing
operational routines \citep{Helfat2007}. Thus, firms with well-functioning DC would be
able to utilize more instances of change and thus more opportunities for learning
(ref). Consequently, high DC leads to more opportunities for both ad-hoc and deliberate
learning to take place and disclose the demand for investments in methods for systematic
experiential learning. As deliberate learning emphasizes a purposefulness of learning by
focusing “on the effect of planned processes for acquiring, codifying, and transferring
existing knowledge within and between organizations” \citep[p. 1160]{Arthur2005}, many
opportunities generated from enhanced DC should lead to more investments in learning
systems. This notion is also supported by previous work emphasizing that doing more of the
same over time leads experience accumulation and hence improved outcomes
\citep{Zollo2004a}. Similarly, by being active in changing its resource base firms
accumulate experience and the importance of deliberate approaches to leading will become
more apparent. A related point is that when the number of opportunities for learning
increases due to higher DC, the incentives for the firm to invest in DL increases due to
potential intensified agency problems \citep{Arthur2005}. The more potential for
knowledge accumulation through learning, the higher potential for misaligned incentives
between workers keeping information back in fear of not being sufficiently compensated,
and management. Hence, the incentive for the firm to invest in DL to better align these
incentives, increases (ibid). In sum these arguments shape an interesting hypothesis on
the feedback loop between DL and DC and how DL could be an outcome as well as an
antecedent of DC \cite{Schilke2018}.(HERE WE ALSO HAVE A DIRECT QUOTE).

\emph{H5: Dynamic capabilities enhances demand for systematic learning so investments in
  deliberate learning abilities will increase over time.}

>>>>

\section{Introduction}

Dynamic capabilities are proposed to confer a competitive advantage by adding unique value to the firm through systematic change, which enhances operational efficiency and enables increased alignment with the environment (Di Stefano et al., 2014; Peteraf et al., 2013)

Not all organizations possess them (Collis, 1996); their path dependencies, intangibility, complexity, and organizational specificity make them hard to imitate (Gibson & Birkinshaw, 2004; Helfat & Winter, 2011); and few other means allow organizations to purposefully change on a continuous basis (Day & Wensley, 1988; Helfat et al., 2007).

Consistent with the theoretical positions of Eisenhardt and Martin (2000), Helfat and Peteraf (2003), Zahra et al. (2006), and Zott (2003), among others, these researchers have argued that dynamic capabilities’ immediate purpose is to change the resource base, and that this change in the resource base, in turn, explains performance variations. According to this argument, resource changes serve as mediators through which dynamic capabilities affect performance. We

One of the major unresolved issues in dynamic capabilities theory is to what extent and
how competitive advantage is the ultimate outcome of dynamic capabilities. This very idea
has

This paper asks two questions: to what extent does a direct lnk between DC and CA exist,
and what determines variations in the role of mediation between them?

The idea that Dc impacts CA is uncontroversial (not). Little is known about how this
happens. The theory holds that this happens through xreating extending and modifying the
resource base. YEt, the change in firm capabilities can themselves lead to CA due to their
evlutionary nature.

Variations in the share of mediation stems from variation sin the environment. When envdyn
is high firms need to act faster and hence tends to choose strategies including resource
orchestration buying in factor markets cause it tkes more time to develop OR.

In less dynamism the role of OR increases and leads to a significant proportion of CA
stemming throhg the mediating effect of OR

There is ongoing interest in (dynamic) capabilities and how they influence firm
performance \citep{Fainshmidt2016a,Teece2014}. Indeed, \emph{dynamic capabilities} (DC),
and its front runner, the resource-based view \citep{Barney2001,Helfat2007}, is now
considered one of the major theories in strategic management
\citep{Schilke2018}. Arguably, the key reason is that this theory clearly articulates how
firms can alter their operating routines through dynamic capabilities. Thus, a key
ingredient in this theory is that DCs can alter firms operating routines, with subsequent
performance implications.

\section{Theory}\label{sec:theory}
Firm capabilities can roughly be divided into two categories. First, ordinary capabilities
(OR) “are directed toward maintaining and leveraging the status quo in terms of the scale
and scope of activities, businesses, product lines, customer segments”
\citep[p. 393]{Schilke2018}. In other words these capabilities are the ones who “permit a
firm to ‘make a living’ in the short term” \citep[p. 991]{Winter2003}. On the other hand,
dynamic capabilities (DC) are “those that operate to extend, modify or create ordinary
capabilities” (ibid). DC theory is emerging as an important paradigm in strategic
management research, but also facing some considerable challenges with respect to its
antecedents and outcomes \citep{Eriksson2014,Schilke2018}, as well as its evolution
and development trajectory \citep{Romme2010,Zollo2002a}.
